From *[P pronoun\scsub{inanim}] to [\tsc{r}-pronoun-P]\\ On \tit{waarmee} in Dutch

### 2020-01-13 paper.tex
einde maandag
vanalles

\title{How P inanimate pronoun becomes R-pronoun-P On waarmee in Dutch}
\author{Fenna Bergsma}
\date{\today}

@@

English has a sentence. This is ungrammatical in Dutch. We need an \tsc{r}-pronoun. The term \tsc{r}-pronoun \citep{riemsdijk1978} refers to a set of nominal elements that can strand prepositions in Dutch (and German). This is noteworthy because Dutch is a normally non-preposition stranding language. \tsc{R}-pronouns morphologically consist of a combination of a preposition and locative pronoun. In the example \ref{ex:klimerop} \tit{op} `on' is the preposition and \tit{er} `there' is the locative pronoun.

@@

\ex.
\ag. *Ik klim op 't.\\
 I climb on it\\
 `I climb on it.'\label{ex:klimopt}
\bg. Ik klim er-op.\\
 I climb there-on\\
 `I climb on it.'\label{ex:klimerop}

@@

This chapter is an in-depth study of the \tsc{r}-pronoun-postposition combination \tit{waarmee} `with what' in Dutch. This instance is interesting for two reasons. First, just like for all \tsc{r}-pronouns, the wh-element is the locative, but there is no meaning component related to location in `with what'. Second, the preposition \tit{met} `with' changes into \tit{mee} `with' when it is combined with an \tsc{r}-pronoun. To my knowledge, there is no account so far for this fact.

@@

This is a free relative construction in which the two predicates (the one in the main clause and the one in the embedded clause) combine with two different cases (i.e. the case requirements do not match). I illustrate this in \ref{ex:gekochtwaarmee}. The predicate in the embedded clause, \tit{schildert} `paint', combines with an instrumental PP. The predicate in the main clause clause, \tit{gekocht} `bought' combines with an accusative DP. The \tsc{r}-pronoun \tit{waarmee} `with what' is used here.\footnote{In this example \tit{mee} `with' is stranded and \tit{waar} `where' is moved to the left edge of the embedded clause. It is also possible for \tit{waar} `where' to bring \tit{mee} `with' along, as in \ref{ex:meealong}, but it is regarded as slightly less natural.

\exg. Ik heb gekocht waar jij mee schildert.\\
 I have bought where you with paint\\
 `I bought what you are painting with.'\label{ex:meealong}
\z.

@@

\exg. Ik heb gekocht waar jij mee schildert.\\
 I have bought where you with paint\\
 `I bought what you are painting with.'\label{ex:gekochtwaarmee}

@@

\ex. \label{ex:objprep}
\ag. Ik schilder samen met haar/hem.\\
 I paint together with her/him\\
 `I am painting together with her/him.'\label{ex:prepani}
\bg. *Ik schilder met 't.\\
 I paint with it\\
 `I am painting with it.'\label{ex:prephet}
\bg. Ik schilder er-mee.\\
 I paint there-with\\
 `I am painting with it.'\label{ex:preper}
\bg. *Ik schilder mee er.\\
 I paint with there\\
 `I am painting with it.'\label{ex:erprep}

@@

\ex.
\ag. Ik zit er-op.\\
 I sit it-on\\
 `I am sitting on it.
\bg. *Ik zit op 't.\\
 I sit on it\\
 `I am sitting on it.

\ex.
 \ag. Hij zwemt er-in.\\
  He swims it-in\\
  `He is swimming in it.'
 \bg. *Hij zwemt in 't.\\
  He swims in it\\
  `He is swimming in it.'

@@

\ex.
\ag. Wat zie jij?\\
 what see you\\
 `What do you see?'\label{ex:wat}
\bg. *Met wat schilder jij?\\
 With what paint you\\
 `What are you painting with?'\label{ex:metwat}
\bg. Waar schilder jij mee?\\
 Where paint you with\\
 `What are you painting with?'\label{ex:waarmee}

@@

\ex.\label{ex:headed}
\ag. Ik schilder met de kwast waar jij ook mee schildert.\\
 I paint with the brush where you also with paint\\
 `I am painting with the brush that you are painting with too.'
\bg. *Ik schilder met de kwast met wat jij ook schildert.\\
 I paint with the brush with what you also paint\\
 `I am painting with the brush that you are painting with too.'

 \ex.\label{ex:headless}
 \ag. Ik schilder waar jij ook mee schildert.\\
  I paint where you also with paint\\
  `I am painting with what you are painting with too.'
 \bg. *Ik schilder met wat jij ook schildert.\\
  I paint with what you also paint\\
  `I am painting with what you are painting with too.'

Table \ref{tbl:inanimates} summarizes the distribution of inanimates in Dutch. Inanimate (personal and wh-)pronouns in Dutch can function as objects verbs, but they are ungrammatical as objects of prepositions. In these contexts, \tsc{r}-pronouns appear.

\begin{table}[ht]
	\center
	\caption {Inanimates in Dutch}
	\begin{minipage}{0.7\linewidth}
		\begin{tabularx}{\textwidth}{ccc}
		\toprule
                        & pers. pronouns & wh-pronouns \\
  	\midrule
objects of verbs        & 't             & wat         \\
objects of prepositions & er             & waar        \\
\bottomrule
\end{tabularx}
\label{tbl:inanimates}
\end{minipage}
\end{table}

@@

\ex.\label{ex:grammatical}
\ag. Ik heb gekocht waar jij mee schildert.\\
 I have bought where you with paint\\
 `I bought what you are painting with.'\label{ex:grammaticalwaar}
\bg. Ik schilder met wat jij hebt gekocht.\\
 I paint with what you have bought\\
 `I paint with what you bought.'\label{ex:grammaticalwat}

@@

\ex.
\ag. Ik schilder er-mee.\\
 I paint there-with\\
 `I am painting with it.'\label{ex:const1}
\bg. Waar schilder jij mee?\\
 Where paint you with\\
 `What are you painting with?'\label{ex:const2}
\bg. Ik schilder met de kwast [waar jij ook mee schildert].\\
 I paint with the brush where you also with paint\\
 `I am painting with the brush that you are painting with too.'\label{ex:const3}
\bg. Ik schilder [waar jij ook mee schildert].\\
 I paint where you also with paint\\
 `I am painting with what you are painting with too.'\label{ex:const4}

@@

\ex.
\a. [[ik] [[schilder] [er-mee]]]\label{ex:const1stage}
\b. [[jij] [[schilder] [[waar] [mee]]]]\label{ex:const2stage}
\b. [[jij] [[ook] [[schilder] [[waar] [mee]]]]]\label{ex:const3stage}

There is additional evidence for the fact that \tit{waarmee} `with what' forms a constituent in the constructions above. In the example in \ref{ex:const2}-\ref{ex:const4} only \tit{waar} `where' was fronted, but \ref{ex:constwaarmee} shows that also the phrase containing \tit{mee} `with' can be moved. As wh-movement can only target constituents, it follows that \tit{waar} `where' and \tit{mee} `with' have to form a constituent.

\ex.\label{ex:constwaarmee}
\ag. Waar-mee schilder jij?\\
 {where-with} paint you\\
 `What are you painting with?'
\bg. Ik schilder [waar-mee jij ook schildert].\\
 I paint {where-with} you also paint\\
 `I am painting with what you are painting with too.'
\bg. Ik schilder met de kwast [waar-mee jij ook schildert].\\
 I paint with the brush {where-with} you also paint\\
 `I am painting with the brush that you are painting with too.'

@@

Let me now show how this applies to the examples with the mismatching free relatives. The two predicates I used in the free relatives are \tit{kopen} `to buy' and \tit{schilderen}  `to paint'. \tit{Kopen} `to buy' takes an accusative DP as its object, illustrated in \ref{ex:kopen}. \tit{Schilderen} `to paint' can take an instrumental PP as its object., shown in \ref{ex:schilderen}.\footnote{\tit{Schilderen} also optionally takes an accusative object, but I am focussing on the instrumental here.}

@@

I repeat the mismatching free relative in which \tit{waarmee} `met wat' appears in \ref{ex:mismatchwaarmee}. The predicate \tit{schildert} `paints' combines in the embedded clause with the instrumental PP. The PP forms a proper constituent within the embedded clause, and the PP can be realized as the \tsc{r}-pronoun \tit{waarmee} `with what'.\footnote{I assume that the embedded clause modifies a phonologically empty noun here which is the DP argument of \tit{gekocht} `bought'.}

\exg. Ik heb gekocht [waar jij mee schildert].\\
 I have bought where you with paint\\
 `I bought what you are painting with.'\label{ex:mismatchwaarmee}

@@

\ex.\label{ex:summaryconst}
\a. *[[met] [wat]] → [waarmee]\label{ex:waarmeefr}
\b. [met [wat [X]]]
\b. [met [[wat] [X]]]\label{ex:metwatfr}

@@

In this section I investigate the internal structure of \tit{waarmee} `with what' and \tit{met wat} `with what' to capture the phonological similarities and differences between the two forms. First, I identify \tit{w} and \tit{a} as morphemes that appear in both expressions. Putting these two aside, I concentrate on \tit{ermee} `with it' and \tit{met 't} `with it'. I decompose these two expressions further, and end up arguing for a picture as shown in Table \ref{tbl:nofeatures}.

The elements \tit{w} and \tit{a} express the same syntactic structure in both expressions. The elements \tit{met} and \tit{'t} together also express the same features as \tit{mee} and \tit{er} together, but the distribution differs. \tit{Met} expresses less structure than \tit{mee}, and \tit{'t} expresses more structure than \tit{er}. \tit{Met} is shown to be an prefix, and \tit{mee} is a suffix. It also becomes clear that the \tit{er} in fact corresponds to the locative \tit{r} in Dutch.

\begin{table}[ht]
	\center
	\caption {\tit{met wat} and \tit{waarmee}}
	\begin{minipage}{0.56\linewidth}
		\begin{tabularx}{\textwidth}{ccccccc}
		\toprule
    \phantom{\tsc{wh}}  & \phantom{\tsc{deix}}                    & \phantom{\tsc{f}4}  & \phantom{\tsc{f}3} & \phantom{\tsc{f}2}  & \phantom{\tsc{f}1}  & \phantom{\tsc{thing}} \\
		%\midrule
    \tit{w}   & \multicolumn{1}{|c|}{\tit{a}}  & \tit{met} & \multicolumn{4}{|c}{\tit{'t}}                \\\hline
    \tit{w}   & \multicolumn{1}{|c|}{\tit{a}}  & \multicolumn{4}{c|}{\tit{mee}}               & \tit{er}  \\
    \bottomrule
\end{tabularx}
\label{tbl:nofeatures}
\end{minipage}
\end{table}

 The idea is that both expressions realize the same features, and \tit{waarmee} `with what' takes precedence when all features form a constituent. The morphemes I distinguish in \tit{waarmee} `with what' are \tit{w}, \tit{-aa-}, \tit{-r} and \tit{mee}. Within \tit{met wat} `with what' I distinguish \tit{met}, \tit{w}, \tit{a} and \tit{-t}. \ref{ex:decompose} shows this as well.

@@

@@

I put \tit{w} and \tit{a} aside for now, assuming they spell out the same syntactic structure in \tit{waarmee} `with what' and \tit{met wat} `with what'. This leaves \tit{er-mee} `with it' and \tit{met 't} `with it'.\footnote{I assume that the \tit{e} in \tit{ermee} `with it' is there for phonological reasons. The vowel is dropped under the right phonological circumstances.

\exg. Ik he niks mee te maken.\\
 I have there nothing with to make\\
 `I have nothing to do with it.'
 \z.

@@

\ex.
\ag. er -mee\\
there with\\
\b. met 't\\
with it\\

I repeat the examples from Section \ref{sec:rdefault} below that show that \tit{ermee} `with it' is used when it forms a proper constituent and \tit{met 't} `with it' is ungrammatical.

\ex.
\ag. Ik schilder er-mee.\\
 I paint there-with\\
 `I am painting with it.'\label{ex:jarmee}
\bg. *Ik schilder met 't.\\
 I paint with it\\
 `I am painting with it.'\label{ex:neemett}

In other words, both expessions realize the same features, and \tit{ermee} `with it' takes precedence when all features form a constituent.

@@

\ex.\label{ex:hemsubobj}
\ag. Hij staat in de hal.\\
 \tsc{3sg.m.nom} stands in the hallway\\
 `He is standing in the hallway.'
\bg. Ik zie hem.\\
 I see \tsc{3sg.m.acc}\\
 `I see it.'
\bg. Ik heb hem een klap gegeven.\\
 I have \tsc{3sg.m.acc} a hit given\\
 `I hit him.'

I assume that the inanimate personal pronoun minimally contains the ontological category \tsc{thing} \citep{kayne2005}. I leave possible number and gender features out because they do not play a role in this paper. For case, I follow \citet{caha2009} that case features case features are organized the containment relation in \ref{ex:casetree}. The higher, more complex cases contain the smaller, less complex cases.

@@

This lexical entry can lexicalize the \tsc{datP}, but also the \tsc{accP} and \tsc{nomP}. This is due to the Superset Principle.

@@

\ex.
\a. \begin{forest} boom
[\tsc{acc}P
    [\tsc{f}2]
    [\tsc{nom}P
        [\tsc{f}1]
        [\tsc{thingP}
            [\tsc{thing}, roof]
        ]
    ]
]
\end{forest}\label{ex:tacc}
\b. \begin{forest} boom
[\tsc{nom}P
    [\tsc{f}1]
    [\tsc{thingP}
        [\tsc{thing}, roof]
    ]
]
\end{forest}\label{ex:tnom}

Let us now turn to \tit{ermee} `with it', which lexicalizes the same structure as \tit{met 't} `with it', but with a different distribution.

\tit{Er} `there' can be used as a locative.

I follow \cite{baunaz2018} in assuming that the ontological category \tsc{location} contains \tsc{thing}.\footnote{\citet{baunaz2018} place in addition \tsc{person} between \tsc{thing} and \tsc{location}, which I left out here.}

 \ex. \begin{forest} boom
 [\tsc{loc}P
 [\tsc{location}]
     [\tsc{thngP}
         [\tsc{thing}, roof]
     ]
 ]
 {\draw (.east) node[right]{⇔ \tit{er}}; }
 \end{forest}\label{ex:entryr}

Notice already here that, via the superset principle, \tit{er} can be used to realize the feature \tsc{thing} as well, as it is contained in \tsc{locP}.

So far, \tit{er} `there' only realizes the feature \tsc{thing} from the structure in \ref{ex:stmetwat}. This leaves \tsc{f1} to \tsc{f4} to be realized by \tit{mee}. I give the lexical tree in \ref{ex:entrymee}.

Note that there are two differences between the lexical entry in \ref{ex:entrymet} and \ref{ex:entrymee}. First, the lexical entry for \tit{mee} `with' in \ref{ex:entrymee}, in addition to the one for \tit{met} `with' in \ref{ex:entrymet}, contains the features \tsc{f1}, \tsc{f2} and \tsc{f3}.
Second, the lexical entry in \ref{ex:entrymet} has a binary bottom, and the lexical entry in \ref{ex:entrymee} has a unary bottom. Nanosyntax distinguishes prefixes and suffixes by distinguishing these types of lexical entries. This is the crucial differences needed for the Dutch instrumental.
\tit{Met} `with' is a prefix (preposition), so it has a binary bottom, and \tit{mee} `with' is a suffix, so it needs a unary bottom. In the next section I show how this works in a derivation.\footnote{For completeness, \tit{mee} `with' has to be blocked from combining with full DPs and animate pronouns (reference). For now I assume that these pronouns spell out an additional feature \tit{σ} that is not realized by \tit{mee} `with' but only by \tit{met} `with'. Full DPs combine in addition with a zero marker for dative case. This way \tit{mee} `with' does not compete with \tit{met} `anymore'. I left this features out in this paper, at is does not contribute to the main point.}\footnote{Giving \tit{met} `with' and \tit{mee} `with' two distinct lexical entries has as a consequence that the phonological overlap between them seems like a coincidence. Moreover, in almost all cases the preposition does not change form when it combines with an \tsc{r}-pronoun, e.g. \tit{in}. If this proposal is on the right track, elements as \tit{in} can be used as either a prefix and as a suffix. A lexical entry as in \ref{ex:presuf} would be a candidate for such an element.

I leave it to future research to determine whether this is a feasible solution.}\footnote{I do not address the ordering difference of the wh-element and the adposition. A topic related to (but not relevant for) this paper is the different positioning of identical adpositions in Dutch. In \ref{ex:dutchin}, \tit{in} changes meaning dependent on whether it proceeds or follows the DP, it is respectively locational or directional.

The different positioning of the adpositions in \tsc{r}-pronouns is driven by spellout and is, therefore, meaningless.}

Table \ref{tbl:withfeatures} gives an overview of the distribution of the features in \tit{met wat} `with what' and \tit{waarmee} `with what'. In both cases, \tit{w} and \tit{a} are the realization of the same morpheme. The difference between the two expressions lies in the distribution of the case features. \tit{Mee} `with' realizes all of them, and \tit{met} `with' only the top two. The residues are realized by respectively \tit{er} `there' and \tit{'t} `it'.

\begin{table}[ht]
	\center
	\caption {\tit{met wat} and \tit{waarmee}}
	\begin{minipage}{0.56\linewidth}
		\begin{tabularx}{\textwidth}{ccccccc}
		\toprule
    \tsc{wh}  & \tsc{deix}                     & \tsc{f}4  & \tsc{f}3 & \tsc{f}2  & \tsc{f}1  & \tsc{thing} \\
		\midrule
    \tit{w}   & \multicolumn{1}{|c|}{\tit{a}}  & \tit{met} & \multicolumn{4}{|c}{\tit{'t}}                \\\hline
    \tit{w}   & \multicolumn{1}{|c|}{\tit{a}}  & \multicolumn{4}{c|}{\tit{mee}}               & \tit{er}  \\
    \bottomrule
\end{tabularx}
\label{tbl:withfeatures}
\end{minipage}
\end{table}

\section{Two mechanisms}

\subsection{Suppletive obliques}

@@

't
't
't
er-van
er-mee
er-mee

\subsection{Prefix or suffix}

https://ling.auf.net/lingbuzz/001026

\ex. mit ein loffel

>>>>

\title{From *[P pronoun\scsub{inanim}] to [\tsc{r}-pronoun-P]\\ On \tit{waarmee} in Dutch}
\author{Fenna Bergsma}
\date{\today}

@@

English has a sentence. This is ungrammatical in Dutch. We need an \tsc{r}-pronoun. The term \tsc{r}-pronoun \citep{riemsdijk1978} refers to a set of nominal elements that can strand prepositions in Dutch (and German). This is noteworthy because Dutch is a normally non-preposition stranding language. \tsc{R}-pronouns morphologically consist of a combination of a preposition and locative pronoun. In the example \ref{ex:klimerop} \tit{op} `on' is the preposition and \tit{'r} `there' is the locative pronoun.

@@

\ex.
\ag. *Ik klim op 't.\\
 I climb on it\\
 `I climb on it.'\label{ex:klimopt}
\bg. Ik klim 'r op.\\
 I climb there on\\
 `I climb on it.'\label{ex:klimerop}

@@

This chapter is an in-depth study of the \tsc{r}-pronoun-postposition combination \tit{waarmee} `with what' in Dutch. This instance is interesting for two reasons. First, just like for all \tsc{r}-pronouns, the wh-element is the locative, but there is no meaning component related to location in `with what'. Second, the preposition \tit{met} `with' changes into \tit{mee} `with' when it is combined with an \tsc{r}-pronoun. This observervation has so far remained unexplained.

@@

This is a free relative construction in which the two predicates (the one in the main clause and the one in the embedded clause) combine with two different cases (i.e. the case requirements do not match). I illustrate this in \ref{ex:gekochtwaarmee}. The predicate in the embedded clause, \tit{schildert} `paint', combines with an instrumental PP. The predicate in the main clause clause, \tit{gekocht} `bought' combines with an accusative DP. The \tsc{r}-pronoun \tit{waarmee} `with what' is used here.\footnote{In this example, \tit{waar} `where' takes \tit{mee} `with' to the left edge of the embedded clause. It is also possible for \tit{mee} `with' to be stranded, and \tit{waar} `where' to be moved to the left edge of the embedded clause on its own.

\exg. Ik heb gekocht waar jij mee schildert.\\
 I have bought waar you with paint\\
 `I bought what you are painting with.'\label{ex:meealong}
\z.

@@

\exg. Ik heb gekocht waarmee jij schildert.\\
 I have bought {where with} you paint\\
 `I bought what you are painting with.'\label{ex:gekochtwaarmee}

@@

\ex. \label{ex:objprep}
\ag. Ik schilder samen met haar/hem.\\
 I paint together with her/him\\
 `I am painting together with her/him.'\label{ex:prepani}
\bg. *Ik schilder met 't.\\
 I paint with it\\
 `I am painting with it.'\label{ex:prephet}
\bg. Ik schilder 'r mee.\\
 I paint there-with\\
 `I am painting with it.'\label{ex:preper}
\bg. *Ik schilder mee 'r.\\
 I paint with there\\
 `I am painting with it.'\label{ex:erprep}

@@

\ex.
\ag. Ik zit 'r op.\\
 I sit there on\\
 `I am sitting on it.
\bg. *Ik zit op 't.\\
 I sit on it\\
 `I am sitting on it.

\ex.
 \ag. Hij zwemt 'r in.\\
  he swims it-in\\
  `He is swimming in it.'
 \bg. *Hij zwemt in 't.\\
  he swims in it\\
  `He is swimming in it.'

@@

\ex.
\ag. Wat zie jij?\\
 what see you\\
 `What do you see?'\label{ex:wat}
\bg. *Met wat schilder jij?\\
 with what paint you\\
 `What are you painting with?'\label{ex:metwat}
\bg. Waarmee schilder jij?\\
 {where with} paint you with\\
 `What are you painting with?'\label{ex:waarmee}

@@

\ex.\label{ex:headed}
\ag. Ik schilder met de kwast waarmee jij ook schildert.\\
 I paint with the brush {where with} you also paint\\
 `I am painting with the brush that you are painting with too.'
\bg. *Ik schilder met de kwast met wat jij ook schildert.\\
 I paint with the brush with what you also paint\\
 `I am painting with the brush that you are painting with too.'

 \ex.\label{ex:headless}
 \ag. Ik schilder waarmee jij ook schildert.\\
  I paint {where with} you also paint\\
  `I am painting with what you are painting with too.'
 \bg. *Ik schilder met wat jij ook schildert.\\
  I paint with what you also paint\\
  `I am painting with what you are painting with too.'

\tit{'t} `it' and \tit{wat} `what' do not combine with prepositions. They are substituted by respectively \tit{'r} `there' and \tit{waar} `where'.

@@

\ex.\label{ex:grammatical}
\ag. Ik heb gekocht waarmee jij schildert.\\
 I have bought {where with} you paint\\
 `I bought what you are painting with.'\label{ex:grammaticalwaar}
\bg. Ik schilder met wat jij hebt gekocht.\\
 I paint with what you have bought\\
 `I paint with what you bought.'\label{ex:grammaticalwat}

@@

\ex.
\ag. Ik schilder 'r mee.\\
 I paint there with\\
 `I am painting with it.'\label{ex:const1}
\bg. Waarmee schilder jij?\\
{where with} paint you with\\
 `What are you painting with?'\label{ex:const2}
\bg. Ik schilder met de kwast [waarmee jij ook schildert].\\
 I paint with the brush {where with} you also paint\\
 `I am painting with the brush that you are painting with too.'\label{ex:const3}
\bg. Ik schilder [waarmee jij ook schildert].\\
 I paint {where with} you also paint\\
 `I am painting with what you are painting with too.'\label{ex:const4}

@@

\ex.
\a. [[ik] [[schilder] ['r mee]]]\label{ex:const1stage}
\b. [[jij] [[schilder] [waarmee]]]\label{ex:const2stage}
\b. [[jij] [[ook] [[schilder] [waarmee]]]]\label{ex:const3stage}

%There is additional evidence for the fact that \tit{waarmee} `with what' forms a constituent in the constructions above. In the example in \ref{ex:const2}-\ref{ex:const4} only \tit{waar} `where' was fronted, but \ref{ex:constwaarmee} shows that also the phrase containing \tit{mee} with' can be moved. As wh-movement can only target constituents, it follows that \tit{waar} `where' and \tit{mee} `with' have to form a constituent.

%\ex.\label{ex:constwaarmee}
%\ag. Waar-mee schilder jij?\\
 %{where-with} paint you\\
 %`What are you painting with?'
%\bg. Ik schilder [waar-mee jij ook schildert].\\
 %I paint {where-with} you also paint\\
 %`I am painting with what you are painting with too.'
%\bg. Ik schilder met de kwast [waar-mee jij ook schildert].\\
 %I paint with the brush {where-with} you also paint\\
 %`I am painting with the brush that you are painting with too.'

@@

Let me now show how this applies to the examples with the mismatching free relatives. The two predicates I used in the free relatives are \tit{kopen} `to buy' and \tit{schilderen}  `to paint'. \tit{Kopen} `to buy' takes an accusative DP as its object, illustrated in \ref{ex:kopen}. \tit{Schilderen} `to paint' can take an instrumental PP as its object., shown in \ref{ex:schilderen}.\footnote{\tit{Schilderen} also optionally takes an (accusative) object, but I am focussing on the instrumental object here.}

@@

I repeat the mismatching free relative in which \tit{waarmee} `met wat' appears in \ref{ex:mismatchwaarmee}. The predicate \tit{schildert} `paints' combines in the embedded clause with the instrumental object. The PP forms a proper constituent within the embedded clause, and the PP can be realized as the \tsc{r}-pronoun \tit{waarmee} `with what'.\footnote{I assume that the embedded clause modifies a phonologically empty noun here which is the DP argument of \tit{gekocht} `bought'.}

\exg. Ik heb gekocht [waarmee jij schildert].\\
 I have bought {where with} you paint\\
 `I bought what you are painting with.'\label{ex:mismatchwaarmee}

@@

\ex.\label{ex:summaryconst}
\a. [[met] [wat]] → [waarmee]\label{ex:waarmeefr}
\b. [met [wat [X]]]
\b. [met [[wat] [X]]]\label{ex:metwatfr}

@@

Along the way I introduce some background

Things to explain
*P pronoun-inanimate

\tit{r} is the locative
\tit{met} changes form

 The idea is that both expressions realize the same featu♦res, and \tit{waarmee} `with what' takes precedence when all features form a constituent. The morphemes I distinguish in \tit{waarmee} `with what' are \tit{w}, \tit{-aa-}, \tit{-r} and \tit{mee}. Within \tit{met wat} `with what' I distinguish \tit{met}, \tit{w}, \tit{a} and \tit{-t}. \ref{ex:decompose} shows this as well.

@@

In this section I investigate the internal structure of \tit{waarmee} `with what' and \tit{met wat} `with what' to capture the phonological similarities and differences between the two forms. First, I identify \tit{w} and \tit{a} as morphemes that appear in both expressions. Putting these two aside, I concentrate on \tit{'rmee} `with it' and \tit{met 't} `with it'. I decompose these two expressions further, and end up arguing for a picture as shown in Table \ref{tbl:nofeatures}.

The elements \tit{w} and \tit{a} express the same syntactic structure in both expressions. The elements \tit{met} and \tit{'t} together also express the same features as \tit{mee} and \tit{'r} together, but the distribution differs. \tit{Met} expresses less structure than \tit{mee}, and \tit{'t} expresses more structure than \tit{'r}. \tit{Met} is shown to be an prefix, and \tit{mee} is a suffix. It also becomes clear that the \tit{'r} in fact corresponds to the locative \tit{r} in Dutch.

\begin{table}[ht]
	\center
	\caption {\tit{met wat} and \tit{waarmee}}
	\begin{minipage}{0.56\linewidth}
		\begin{tabularx}{\textwidth}{ccccccc}
		\toprule
    \phantom{\tsc{wh}}  & \phantom{\tsc{deix}}                    & \phantom{\tsc{f}4}  & \phantom{\tsc{f}3} & \phantom{\tsc{f}2}  & \phantom{\tsc{f}1}  & \phantom{\tsc{thing}} \\
		%\midrule
    \tit{w}   & \multicolumn{1}{|c|}{\tit{a}}  & \tit{met} & \multicolumn{4}{|c}{\tit{'t}}                \\\hline
    \tit{w}   & \multicolumn{1}{|c|}{\tit{a}}  & \multicolumn{4}{c|}{\tit{mee}}               & \tit{'r}  \\
    \bottomrule
\end{tabularx}
\label{tbl:nofeatures}
\end{minipage}
\end{table}

\subsection{Overlap: \tit{w-} and \tit{-a-}}

@@

I put \tit{w} and \tit{a} aside for now, assuming they spell out the same syntactic structure in \tit{waarmee} `with what' and \tit{met wat} `with what'. This leaves \tit{'r-mee} `with it' and \tit{met 't} `with it'.\footnote{I assume that the \tit{e} in \tit{'rmee} `with it' is there for phonological reasons. The vowel is dropped under the right phonological circumstances.

\exg. Ik heb 'r niks mee te maken.\\
 I have there nothing with to make\\
 `I have nothing to do with it.'
 \z.

@@

\ex.
\ag. 'r -mee\\
there with\\
\b. met 't\\
with it\\

\subsection{Differences: \tit{'r} vs. \tit{'t} and \tit{met} vs. \tit{-mee}}

Having put \tit{w-} and \tit{a} aside for now, the forms \tit{met 't} `with it` and \tit{'rmee} `with it' are left. Below I repeat examples from Section \ref{sec:rdefault} that show that \tit{met} `with` can express instrumental case, and \tit{'t} is an inanimate pronoun. \ref{ex:metdp} shows \tit{met} `with' combined with full DP. \ref{ex:tverb} shows \tit{'t} as the object of verbs.

\ex.
\ag. Ik schilder met een kwast.\\
 I paint with a brush\\
 `I am painting with a brush.'\label{ex:metdp}
 \bg. Ik zie 't.\\
  I see it\\
  `I see it.'\label{ex:tverb}

However, \tit{met} `with' and \tit{'t} `it' do not appear together if they form a proper constituent. Instead, \tit{'rmee} is used, as shown in \ref{ex:rmeeconst}.

\ex.\label{ex:rmeeconst}
\ag. Ik schilder 'r-mee.\\
 I paint there-with\\
 `I am painting with it.'\label{ex:jarmee}
\bg. *Ik schilder met 't.\\
 I paint with it\\
 `I am painting with it.'\label{ex:neemett}

In this section I set up an account that makes the ungrammaticality of \tit{met 't} and the appearance of \tit{'rmee} follow from spellout. The analysis accounts for the following two observations, taking \tit{met 't} as the point of departure. First, \tit{met} `with' changes from being a preposition to being a suffix, and its form changes into \tit{mee}. This process is restricted to inanimate pronouns, and it does not apply to full DPs and animate pronouns. Second, \tit{'t} is replaced by \tit{'r}, a morpheme that is associated with the locative in Dutch.

This section first discusses some basic facts about case and assumptions about Nanosyntax. Then I discuss the distinction between prefixes and suffixes, followed by a discussion on suppletion in pronouns. Finally, I specify the lexical entries for \tit{'rmee} `with it` and \tit{met 't} `with it` that account for the described facts.

\subsubsection{Case and nanosyntax}

@@

\ex.\label{ex:hemsubobj}
\ag. Hij staat in de hal.\\
 \tsc{3sg.m.nom} stands in the hallway\\
 `He is standing in the hallway.'
\bg. Ik zie hem.\\
 I see \tsc{3sg.m.acc}\\
 `I see it.'
\bg. Ik heb hem een klap gegeven.\\
 I have \tsc{3sg.m.dat} a hit given\\
 `I gave him a smack.'

I assume that the inanimate personal pronoun minimally contains the ontological category \tsc{thing} \citep{kayne2005}. I leave possible number and gender features out because they do not play a role in this paper.

@@

 IT CANNOT BE STRONG, but it is weak 't staat in de hal

This lexical entry can lexicalize the \tsc{datP}, but also the \tsc{accP} and \tsc{nomP}, due to the Superset Principle.

START WITH 'T IN DUTCH, JUST HOW I DID BEFORE. SUPERSET PRINCIPLE, CASE HIERARCHY, ALL OF IT

THEN INTRODUCE PREFIX/SUFFIX DISTINCTION.

I discuss the basics of spellout and case..

\exg. mit ein-em Löffel\\
with a-\tsc{dat.sg} spoon\\
`with a spoon' \hfill (German)

\ex. The preposition/suffix hierarchy
\a. If the expression of a particular case in the Case sequence (below) involves a preposition, then all cases to its right do as well.
\b. The Case sequence: \tsc{nom – acc – gen – dat – ins – com} \hfill \citep{2009}

put this in syntax

now derive it with movements

Spellout happens in a cyclic derivation, following a spellout algorithm \citep{starke2018}. After each instance of merge, spellout takes place. If no spellout exist for the phrase created by the newly added feature, evacuation movements specified in the spellout algorithm take place. The algorithm is given in \ref{ex:spellout}.

\ex.
\a. \begin{forest} boom
[\tsc{ins{}}
    [\tsc{f5}, roof]
]
{\draw (.east) node[right]{⇔ \tit{mit}}; }
\end{forest}\label{ex:mit}
\b. \begin{forest} boom
[\tsc{dat}P
    [\tsc{f}3]
    [\tsc{acc}P
        [\tsc{f}2]
        [\tsc{nom}P
            [\tsc{f}1]
        ]
    ]
]
{\draw (.east) node[right]{⇔ \tit{-em}}; }
\end{forest}\label{ex:em}

If not match can be found, a specifier is constructed.

For case, I follow \citet{caha2009} that case features case features are organized the containment relation in \ref{ex:casetree}. The higher, more complex cases contain the smaller, less complex cases.

@@

Nanosyntax distinguishes prefixes and suffixes by distinguishing these types of lexical entries. This is the crucial difference needed for the Dutch instrumental.

The same intuition will hold for Dutch.
-mee is going to be a suffix that will be able to attach to 'r, but it will not to full DPs and animates.

\subsubsection{Backtracking}

@@

't
't
't
'r-van
'r-mee
'r-mee

\subsubsection{The lexical entries}

Let us now turn to \tit{'rmee} `with it', which lexicalizes the same structure as \tit{met 't} `with it', but with a different distribution.

\tit{'r} `there' can be used as a locative.

I follow \cite{baunaz2018} in assuming that the ontological category \tsc{location} contains \tsc{thing}.

\footnote{\citet{baunaz2018} place in addition \tsc{person} between \tsc{thing} and \tsc{location}, which I left out here.}

 \ex. \begin{forest} boom
 [\tsc{loc}P
 [\tsc{location}]
     [\tsc{thingP}
         [\tsc{thing}, roof]
     ]
 ]
 {\draw (.east) node[right]{⇔ \tit{'r}}; }
 \end{forest}\label{ex:entryr}

Notice already here that, via the superset principle, \tit{'r} can be used to realize the feature \tsc{thing} as well, as it is contained in \tsc{locP}.

So far, \tit{'r} `there' only realizes the feature \tsc{thing} from the structure in \ref{ex:stmetwat}. This leaves \tsc{f1} to \tsc{f4} to be realized by \tit{mee}. I give the lexical tree in \ref{ex:entrymee}.

Note that there are two differences between the lexical entry in \ref{ex:entrymet} and \ref{ex:entrymee}. First, the lexical entry for \tit{mee} `with' in \ref{ex:entrymee}, in addition to the one for \tit{met} `with' in \ref{ex:entrymet}, contains the features \tsc{f1}, \tsc{f2} and \tsc{f3}.
Second, the lexical entry in \ref{ex:entrymet} has a binary bottom, and the lexical entry in \ref{ex:entrymee} has a unary bottom. \tit{Met} `with' is a prefix (preposition), so it has a binary bottom, and \tit{mee} `with' is a suffix, so it needs a unary bottom. In the next section I show how this works in a derivation.

\footnote{For completeness, \tit{mee} `with' has to be blocked from combining with full DPs and animate pronouns (reference). For now I assume that these pronouns spell out an additional feature \tit{σ} that is not realized by \tit{mee} `with' but only by \tit{met} `with'. Full DPs combine in addition with a zero marker for dative case. This way \tit{mee} `with' does not compete with \tit{met} `anymore'. I left this features out in this paper, at is does not contribute to the main point.}
